% DES-LOC Appendix - Reconstructed from PDF
% Claude-1 (M001-M025): Key appendix sections
% Appendices A, C, F, G, H, I, J, K, L, M
% (Appendices B, D, E contain extensive figures/proofs → deferred to Claude 9-14)

\appendix

%=============================================================================
\section{Experimental Details and Optimizer Hyperparameter Sweeps}
\label{sec:appendix_a}
%=============================================================================

Here we provide additional experimental details complementing those in Section~4.1, including: a) model architecture details and hyperparameters independent of optimizer choice (Section~\ref{sec:arch_details}), b) our hyperparameter sweep procedure to select optimizer-specific settings (Section~\ref{sec:sweep}), and c) the optimal hyperparameters with those used in Section~5 highlighted in bold.

\subsection{Architecture Details and Hyperparameters}
\label{sec:arch_details}

\begin{table}[h]
\centering
\caption{Model architecture and training parameters. We denote the number of transformer blocks by \#Blocks, number of attention heads by \#Heads, embedding dimension by $d_{\text{model}}$, vocabulary size by $|V|$, and feedforward-layer expansion by Exp.\ Ratio. All models use positional embeddings~\citep{su2024roformer}, the silu activation function, and norm-based gradient clipping with clip-bound $\rho$. Global batch size (summed across all workers) is $|B_G|$, and sequence length is standard for models at these scales. For model initialization we use $\sigma = 1/\sqrt{d_{\text{model}}}$. The total number of steps is denoted by $T$.}
\label{tab:arch}
\begin{tabular}{lccccccccccc}
\toprule
Model Size & Blocks & $d_{\text{model}}$ & $|V|$ & \#Heads & Exp.\ Ratio & ROPE $\theta$ & ACT & Init $\sigma$ & $\rho$ & Seq Len & $|B_G|$ & $T$ \\
\midrule
135M & 30 & 576  & 50K & 9  & 4 & 10000 & silu & 0.04 & 1.0 & 2048 & 1024 & 1536, 3072 \\
720M & 12 & 2048 & 50K & 16 & 4 & 10000 & SiLU & 0.02 & 1.0 & 2048 & 512  & 38912 \\
1.7B & 24 & 2048 & 50K & 16 & 4 & 10000 & silu & 0.02 & 1.0 & 2048 & 1024 & 20480 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:arch} summarizes the architectural details of our models, following established practices for large
language models at their respective scales. Unless otherwise stated, we adopt the hyperparameters
recommended by~\citet{allal2025smollm2} for both the 135M and the 1.7B models. We operate at a
batch size of 2M tokens, which is very large for the 135M model at the length of training we
perform~\citep{zhang2025critical} and industry-standard for the 1.7B model~\citep{touvron2023llama2}, we
chose to operate at large batch sizes because adaptive optimizers provide benefits primarily in large-batch training regimes~\citep{kunstner2023noise}. Moreover, we intend \texttt{DES-LOC} for use in cross data-center scenarios, where effectively utilizing available accelerators naturally demands large batch sizes and/or model scales. For both model sizes, we train for approximately $2\times$ the compute-optimal token budget~\citep{hoffmann2022training}, placing our evaluations within the context of extended-duration foundation model training~\citep{allal2025smollm2}. Our chosen token budget is conservative due to resource constraints; for comparison, \citet{allal2025smollm2} used 11 trillion tokens which is over $4000\times$ compute-optimal for the 135M model, and $300\times$ for the 1.7B.

We select warmup and decay schedules following recommendations from~\citet{zhang2025critical,hagele2024scaling,allal2025smollm2}. For the 135M model, the warmup period is set to $T_{\text{WARM}} = 512$
steps, corresponding to roughly 40\% of the compute-optimal training tokens recommended by~\citet{zhang2025critical}. For the 1.7B model, we use the recommended $T_{\text{WARM}} = 2048$ steps from~\citet{allal2025smollm2}, roughly 10\% of total training. The stable-decay period uses a $1 - \text{SQRT}$ schedule over
the final $T_{\text{DECAY}} = 10\% \times T$ steps~\citep{hagele2024scaling}. For shorter runs, such as $T = 1536$ during heterogeneous-data evaluations, we keep the warmup fixed and proportionally scale the decay to ensure well-conditioned parameter updates during the stable learning rate period. The seeds we use for data sampling and for controlling the training algorithms and model are provided in the code accompanying the appendix.

\subsection{Optimizer Parameters Sweeping Procedure}
\label{sec:sweep}

As detailed in Section~\ref{sec:desloc} and verified empirically in Section~\ref{sec:rq2}, the choice of decay rates $\beta_1, \beta_2$
strongly influences the effective synchronization frequencies achievable by both DES-LOC and
Local Adam. This relationship arises directly from the half-life of optimizer states, given by
$\tau_{0.5} = \frac{\ln(0.5)}{\ln(\beta)}$.

For Adam, prior studies such as \citet{wortsman2023stable} have demonstrated a critical interplay between the learning rate ($\eta$), batch size, and the second-momentum decay $\beta_2$. Specifically, increasing either the learning rate or batch size typically demands a lower $\beta_2$ to maintain training stability and avoid loss spikes. Conversely, higher $\beta_2$ values constrain the learning rate and batch size. Such dynamics have also been recently observed between the learning rate and the first-momentum decay $\beta_1$ in~\citet{pagliardini2025ademamix}. Given that all our experiments use a fixed large batch size of roughly 2 million tokens (appropriate for billion-scale training), we systematically tune the learning rate $\eta$ in response to changes in $\beta_1, \beta_2$. We try values of $\beta_1, \beta_2$ based on previous works~\citep{zhang2025critical} and follow the theoretical convergence requirement of~\citet{zhang2022adam} setting $\beta_1 \leq \sqrt{\beta_2}$.

Due to computational constraints, we cannot jointly optimize synchronization periods, data distributions, and decay parameters, and instead adopt a structured two-stage tuning approach:
\begin{enumerate}
    \item \textbf{Stage 1: Tuning $\eta$ for DDP.} Starting from the recommended baseline learning rate ($\eta_0$)
    from~\citet{allal2025smollm2}, we conduct a grid search as outlined by~\citet{charles2025communication}:
    $\{\ldots, \sqrt{2}^{-2}\eta_0, \sqrt{2}^{-1}\eta_0, \eta_0, \sqrt{2}\eta_0, \sqrt{2}^2\eta_0, \ldots\}$
    We expand this search until perplexity stops improving, identifying an optimal learning rate $\eta^*_{\text{DDP}}$ for each $(\beta_1, \beta_2)$ configuration.

    \item \textbf{Stage 2: Tuning $\eta$ for Local Adam.} We then repeat this procedure for Local Adam,
    using $\eta^*_{\text{DDP}}$ as the new baseline. To balance generalizability and computational cost, we set
    the synchronization period to an intermediate value of $K = 64$, between high-frequency ($K = 16$) and low-frequency ($K = 256$) scenarios.
\end{enumerate}

Additionally, following~\citet{zhang2025critical}, we omit weight decay (set to zero) to simplify the hyperparameter tuning process, as it directly affects only model parameters, not optimizer states.

For experiments using Nesterov, we follow the hyperparameter sweeping procedure of~\citet{charles2025communication}, starting with a server learning rate of $1.0$ and a momentum of $0.9$ and only lowering it if it fails to converge.

\subsubsection{Optimizers' Hyperparameter Configurations}
\label{sec:hp_configs}

\begin{table}[h]
\centering
\caption{Optimal learning rates $\eta^*$ for $\beta_1, \beta_2$ configurations of ADOPT/Adam. The hyperparameter sweep procedure (see Section~\ref{sec:sweep}) involves incrementally adjusting the learning rate by factors of $\sqrt{2}$ around the initial value from~\citet{allal2025smollm2} until performance stops improving.}
\label{tab:lr_sweep}
\begin{tabular}{llll}
\toprule
Optimizer & $\beta_1$ & $\beta_2$ & $\eta^*$ \\
\midrule
\multirow{4}{*}{ADOPT} & 0.9   & 0.9999 & 0.0021 \\
                        & \textbf{0.95}  & \textbf{0.9999} & \textbf{0.0021} \\
                        & 0.99  & 0.9999 & 0.0014 \\
                        & 0.995 & 0.9999 & 0.0007 \\
\midrule
\multirow{5}{*}{Adam}  & 0.9   & 0.95   & 0.0042 \\
                        & \textbf{0.95}  & \textbf{0.95}   & \textbf{0.003}  \\
                        & 0.9   & 0.99   & 0.003  \\
                        & 0.95  & 0.99   & 0.003  \\
                        & 0.99  & 0.99   & 0.0021 \\
\bottomrule
\end{tabular}
\end{table}

Our hyperparameter sweep (Table~\ref{tab:lr_sweep}) indicates that the optimal learning rate $\eta^*$ under the warmup-stable-decay scheduler~\citep{hagele2024scaling} strongly depends on both optimizer type and the chosen $\beta_1, \beta_2$ values. For Adam, optimal learning rates and second-momentum decay ($\beta_2$) align closely with recommendations from~\citet{allal2025smollm2}, though a slightly higher first-momentum decay ($\beta_1$) consistently performs better, in agreement with prior findings~\citep{zhang2025critical}. For ADOPT (default $\beta_2$), we observe a lower optimal learning rate compared to Adam, but similar best-performing $\beta_1$ values. We also find that the optimal learning rate does not differ between DDP and Local Adam for given $\beta_1, \beta_2$ when $K = 64$ and using a $\sqrt{2}$ sweep, higher learning rates either do not provide a benefit or diverge while lower learning rates are only necessary when pushing $K$ far closer to the complete training duration.

We find that increasing $\beta_1$ for ADOPT, and $\beta_1, \beta_2$ for Adam, leads to rapid performance degradation, particularly at or above $0.99$. Since the half-life at $\beta = 0.99$ ($\tau_{0.5} \approx 69$) is not sufficiently longer than at $\beta = 0.95$ ($\tau_{0.5} \approx 13.5$) to justify the observed performance drop, we select $\beta_1 = 0.95$ for all experiments, along with the default $\beta_2$ for ADOPT and $\beta_2 = 0.95$ for Adam.

\paragraph{Takeaway.} Increasing an optimizer state's $\beta$ significantly affects performance. Since linear increases in $\beta$ cause only logarithmic changes in half-life $\tau_{0.5}$, raising $\beta$ beyond the optimal value degrades performance without substantially improving the achievable synchronization frequency (Section~\ref{sec:rq2}).

%=============================================================================
\section{Complementary Results to Sections 2.1 and 5}
\label{sec:complementary}
%=============================================================================

% NOTE: This section contains extensive figures (Figs 8-22) and experimental results.
% Full LaTeX reconstruction deferred to Claude instances 9-10 (M201-M250).
% Key takeaways summarized here for completeness.

% B.1: Toy Problem on Non-IID Data → DES-LOC converges, heuristics fail
% B.2: Independent sync frequency ablations → confirms theory
% B.3: Communication reduction baselines → 2x over Local Adam
% B.4: FAVG+OPT instabilities → activation norm growth
% B.5: DES-LOC Nesterov vs DiLoCo at 1B → DES-LOC competitive
% B.6: Very low bandwidth (10 Gbit/s) → DES-LOC still beneficial
% B.7: Muon experimental details → Kx=32, Ku=96
% B.8: Flux vision model → DES-LOC generalizes beyond LLMs
% B.9: 7B throughput on B200 → near-linear scaling

%=============================================================================
\section{Further Algorithmic Details of DES-LOC}
\label{sec:appendix_c}
%=============================================================================

\subsection{Extension to FedOpt}
\label{sec:fedopt}

Although~\citet{cheng2025convergence} show provable convergence for adaptive inner optimizers in a
federated optimization framework, their result rests on the assumption that after a period of local
work, the new global model is created by averaging the local client models. In relation to the larger
FedOpt literature~\citep{reddi2021adaptive}, the scheme chosen by~\citet{cheng2025convergence} resembles
that of FedAvg, or where the server optimizer is SGD with the outer learning rate set to one.

We argue that indeed DES-LOC's principles can be effectively applied to any FedOpt method and
not just FedAvg. While using an alternate server optimizer does not have proven convergence
guarantees as yet, we show in Algorithm~\ref{alg:desloc} that the choice of the ServerOpt is not constrained
from a practical point-of-view. However, the improvements that DES-LOC provide are related to
the local optimization procedure, which is orthogonal to the outer optimizer choice.

\subsection{Deterministic Optimizer-specific Variants of Algorithm 1}
\label{sec:algorithm_variants}

\begin{algorithm}[h]
\caption{DES-LOC-Adam}
\label{alg:desloc_adam}
\begin{algorithmic}[1]
\REQUIRE Model tensors, Hyper-parameters
\STATE $x_0, u_{-1}, v_{-1} \in \mathbb{R}^d$ \COMMENT{initial parameter vector, seeds for first and second moments}
\STATE $\{\eta_t\}_{t=0}^{T-1} \subset \mathbb{R}_{>0}$ \COMMENT{step-size schedule}
\STATE $\beta_1, \beta_2 \in [0,1)$ \COMMENT{Adam decay factors}
\STATE $\rho, \lambda \in \mathbb{R}_{>0}$ \COMMENT{gradient clipping term, $\ell_2$ stability term}
\STATE $T, M \in \mathbb{N}_+$ \COMMENT{total iterations, number of workers}
\STATE $K_x, K_u, K_v \in \mathbb{N}_+$ \COMMENT{sync periods for parameters, first and second moments}
\ENSURE $x_T, u_{T-1}, v_{T-1}$
\STATE for each worker $m$: $x^m_0 = x_0$, $u^m_{-1} = v^m_{-1} = 0$
\FOR{$t = 0, \ldots, T-1$}
    \FOR{all workers $m = 0, \ldots, M-1$ \textbf{in parallel}}
        \STATE $g^m_t \gets \nabla F(x^m_t; \xi^m_t)$
        \STATE $\hat{g}^m_t \gets \text{clip}(g^m_t, \rho)$
        \IF{$t \bmod K_u = 0$}
            \STATE $u^m_t \gets \beta_1 \mathbb{E}_m[u^m_{t-1}] + (1 - \beta_1)\hat{g}^m_t$
        \ELSE
            \STATE $u^m_t \gets \beta_1 u^m_{t-1} + (1 - \beta_1)\hat{g}^m_t$
        \ENDIF
        \IF{$t \bmod K_v = 0$}
            \STATE $v^m_t \gets \beta_2 \mathbb{E}_m[v^m_{t-1}] + (1 - \beta_2)(\hat{g}^m_t \odot \hat{g}^m_t)$
        \ELSE
            \STATE $v^m_t \gets \beta_2 v^m_{t-1} + (1 - \beta_2)(\hat{g}^m_t \odot \hat{g}^m_t)$
        \ENDIF
        \STATE $m^m_t \gets \beta_1 m^m_{t-1} + (1 - \beta_1) \frac{\hat{g}^m_t}{\max\{\sqrt{v^m_{t-1}}, \epsilon\}}$ \COMMENT{ADOPT update}
        \STATE $d^m_t \gets \eta_t m^m_t$
        \IF{$t \bmod K_x = 0$}
            \STATE $x^m_{t+1} \gets \mathbb{E}_m[x^m_t] - d^m_t$
        \ELSE
            \STATE $x^m_{t+1} \gets x^m_t - d^m_t$
        \ENDIF
    \ENDFOR
\ENDFOR
\end{algorithmic}
\end{algorithm}

%=============================================================================
% Appendix D: Convergence Analysis of DES-LOC-SGDM
% NOTE: Full 8-page proof deferred to Claude instances 12-13 (M276-M325)
%=============================================================================
\section{Convergence Analysis of DES-LOC-SGDM (in expectation bounds)}
\label{sec:sgdm_proof}
% Placeholder — full proof spans pages 32-43 of the PDF

%=============================================================================
% Appendix E: High-probability bounds for DES-LOC-Adam
% NOTE: Full proof deferred to Claude instance 14 (M326-M350)
%=============================================================================
\section{Convergence Analysis of DES-LOC-Adam (high-probability bounds)}
\label{sec:high_prob}
% Placeholder — full proof spans pages 44-46 of the PDF

%=============================================================================
\section{Derivation of Eqs.\ (1) and (2): Maximum Momentum Change With Clipping}
\label{sec:drift_derivation}
%=============================================================================

\textbf{Step 2: Bound on $\|u_{t+K} - u_t\|_\infty$.} Now we bound the change in the momentum over $K$ steps
explicitly. Unrolling the recursion, we have:
\begin{equation}
u_{t+K} = \beta_1^K u_t + (1 - \beta_1) \sum_{k=0}^{K-1} \beta_1^k g_{t+K-k}.
\label{eq:unroll_u}
\end{equation}

Subtracting $u_t$ from both sides, we obtain:
\begin{equation}
u_{t+K} - u_t = (\beta_1^K - 1)u_t + (1 - \beta_1) \sum_{k=0}^{K-1} \beta_1^k g_{t+K-k}.
\label{eq:diff_u}
\end{equation}

Applying the triangle inequality gives:
\begin{equation}
\|u_{t+K} - u_t\|_\infty \leq |1 - \beta_1^K| \|u_t\|_\infty + (1 - \beta_1) \sum_{k=0}^{K-1} \beta_1^k \|g_{t+K-k}\|_\infty.
\label{eq:triangle_u}
\end{equation}

Using $\|g_{t,i}\|_\infty \leq \rho$ (from coordinate-wise clipping) and $\|u_t\|_\infty \leq \rho$ (by induction, since $u$ is an exponential moving average of clipped gradients), the geometric series evaluates to:
\begin{equation}
\|u_{t+K} - u_t\|_\infty \leq (1 - \beta_1^K)\rho + (1 - \beta_1^K)\rho = 2\rho(1 - \beta_1^K).
\end{equation}

The identical argument applies to the second moment $v_t$ with $\rho$ replaced by $\rho^2$ (since $(g_t \odot g_t)_i \leq \rho^2$), yielding:
\begin{equation}
\|v_{t+K} - v_t\|_\infty \leq 2\rho^2(1 - \beta_2^K).
\end{equation}

%=============================================================================
\section{Wall-Clock Time Modeling}
\label{sec:wallclock_model}
%=============================================================================

\subsection{Estimating Total Wall-Clock Time}

We model the total wall-clock time $\mathcal{T}$ for training as:
\begin{equation}
\mathcal{T} = T \cdot t_{\text{step}} + R_x \cdot t_{\text{comm}}(x) + \sum_{j=1}^{N} R_j \cdot t_{\text{comm}}(s^j),
\end{equation}
where $T$ is the total number of steps, $t_{\text{step}}$ is the per-step computation time, $R_x = T/K_x$ is the number of parameter sync rounds, $R_j = T/K_j$ is the number of sync rounds for state $j$, and $t_{\text{comm}}(\cdot)$ is the communication time for a given tensor.

For Ring-AllReduce with model size $|x|$ parameters, bandwidth $B_w$ (bytes/sec), and $M$ workers:
\begin{equation}
t_{\text{comm}}(x) = \frac{2(M-1)}{M} \cdot \frac{|x| \cdot \text{bytes\_per\_param}}{B_w}.
\end{equation}

DES-LOC's communication cost relative to DDP is:
\begin{equation}
\frac{\mathcal{T}_{\text{DES-LOC}}}{\mathcal{T}_{\text{DDP}}} = \frac{T \cdot t_{\text{step}} + \frac{T}{K_x} t_{\text{comm}}(x) + \frac{T}{K_u} t_{\text{comm}}(u) + \frac{T}{K_v} t_{\text{comm}}(v)}{T \cdot (t_{\text{step}} + t_{\text{comm}}(x) + t_{\text{comm}}(u) + t_{\text{comm}}(v))}.
\end{equation}

With $K_x = 256$, $K_u = 3K_x = 768$, $K_v = 6K_x = 1536$, DES-LOC communicates $\approx 170\times$ less than DDP.

\begin{table}[h]
\centering
\caption{Throughput (tokens/sec/GPU) at different scales. DES-LOC matches FAVG+OPT throughput.}
\label{tab:throughput}
\begin{tabular}{l|cc|cc}
\toprule
& \multicolumn{2}{c|}{1B Model} & \multicolumn{2}{c}{7B Model} \\
$K_x$ & 16 & 256 & 16 & 256 \\
\midrule
DDP          & 60k & 60k & 8.5k & 8.5k \\
FAVG+OPT     & 106k & 135k & 11.5k & 13.7k \\
Local Adam   & 89k & 132k & 10.4k & 13.6k \\
DES-LOC      & 105k & 134k & 11.4k & 13.7k \\
\bottomrule
\end{tabular}
\end{table}

%=============================================================================
\section{Checkpointing vs.\ Periodic State Synchronization}
\label{sec:checkpoint}
%=============================================================================

A natural question is whether periodic checkpointing can replace DES-LOC's explicit state synchronization for handling worker failures. We argue it cannot, for two reasons:

\begin{enumerate}
\item \textbf{State staleness.} When a new worker joins from a checkpoint, its optimizer states ($u, v$) reflect the global state at checkpoint time. If the checkpoint is old, these states are stale and can cause training instability. DES-LOC's periodic synchronization ensures all workers' states stay within bounded drift (Eqs.~\ref{eq:drift_u},~\ref{eq:drift_v}).

\item \textbf{No convergence guarantees.} Ad-hoc checkpoint averaging provides no theoretical guarantees about the quality of the resulting optimizer states. DES-LOC's provable convergence (Theorem~\ref{thm:sgdm}) holds precisely because synchronization is periodic and bounded.
\end{enumerate}

%=============================================================================
\section{Choosing Synchronization Frequencies}
\label{sec:choosing_freq}
%=============================================================================

For a state with a decay rate $\beta$ approaching 1, the half-life can be calculated as:
\begin{equation}
t_{1/2} \approx \frac{\ln 2}{1 - \beta}.
\end{equation}

Based on this, we propose the following two-step methodology:

\paragraph{Parameters First ($K_x$).} The synchronization of model parameters is paramount to training
quality. The period $K_x$ should be chosen to match the end-of-training quality of fully-synchronous DDP at a target step budget while still materially reducing communication. In
practice, starting points like $K_x = 16$ or $K_x = 32$ are effective.

\paragraph{Momentum by Half-Life ($K_u, K_v$).} For any optimizer momentum state with a decay rate
$\beta$, its synchronization period $K$ should be set near its calculated half-life, i.e., $K \approx t_{1/2}$.
For common optimizers like Adam or ADOPT with well-tuned decay rates $\beta_1$ and $\beta_2$, this
simplifies to:
\begin{equation}
K_u \approx \frac{\ln 2}{1 - \beta_1}, \quad K_v \approx \frac{\ln 2}{1 - \beta_2}.
\end{equation}

Following this heuristic can yield a minimum $5\times$ reduction in communication cost over DDP for
$K_x \geq 16$, significantly decreasing wall-clock time while achieving convergence speed and final
model quality comparable to DDP.

%=============================================================================
\section{Critical Batch Size and Regime Positioning}
\label{sec:critical_batch}
%=============================================================================

To formally contextualize the regimes where DES-LOC is most beneficial, we begin with the statistical
properties of gradient estimation. Let the true gradient over the full data distribution for a loss function
$L: \mathbb{R}^d \to \mathbb{R}$ be $G(\theta) = \nabla L(\theta)$. In practice, a mini-batch of size $B$ provides an estimate, $G_{\text{est}}(\theta)$.
The variance of this estimator scales inversely with the batch size:
\begin{equation}
\text{Var}[G_{\text{est}}(\theta)] = \frac{1}{B}\Sigma(\theta),
\end{equation}
where $\Sigma(\theta)$ is the per-sample gradient covariance. DES-LOC is most beneficial in the large-batch regime where communication costs dominate compute costs.

%=============================================================================
\section{Extended Related Work}
\label{sec:extended_related}
%=============================================================================

\paragraph{Gradient compression and sparsification.}
Methods like Deep Gradient Compression~\citep{lin2018deep}, TopK sparsification~\citep{alistarh2018convergence}, and CocktailSGD~\citep{wang2023cocktailsgd} reduce communication by compressing or sparsifying gradients. These are orthogonal to DES-LOC and can be combined for further savings.

\paragraph{Model parallelism.}
Megatron-LM~\citep{shoeybi2019megatron}, ZeRO~\citep{rajbhandari2020zero}, and FSDP~\citep{zhao2023pytorch} partition model parameters across devices. DES-LOC focuses on data parallelism and is complementary to model-parallel approaches.

\paragraph{Personalized and heterogeneous federated learning.}
Methods like Ditto~\citep{li2021ditto}, FedProx~\citep{li2020federated}, and Per-FedAvg~\citep{fallah2020personalized} address non-IID data in federated settings. DES-LOC inherits robustness to heterogeneity from its theoretical framework (Assumption~\ref{asm:heterogeneity}).

%=============================================================================
\section{LLM Usage Declaration}
\label{sec:llm_usage}
%=============================================================================

Large Language Models were used during the preparation of this work for editing and improving the
text quality of the manuscript. No LLM was used for generating scientific content, proofs, or
experimental results. All technical contributions are the original work of the authors.

%=============================================================================
\section{Limitations}
\label{sec:limitations}
%=============================================================================

\begin{enumerate}
\item \textbf{Communication overlap.} Our current implementation uses a ``stop-the-world'' synchronization approach. Overlapping communication with computation~\citep{douillard2025streaming} could further improve wall-clock time.

\item \textbf{Heterogeneous hardware.} We assume homogeneous worker hardware. Extending DES-LOC to heterogeneous clusters with different compute speeds is left for future work.

\item \textbf{Adaptive synchronization.} Our synchronization periods are fixed throughout training. Adapting them based on training dynamics (e.g., increasing $K_x$ as training progresses) could yield further improvements.

\item \textbf{Scale of experiments.} While we validate up to 1.7B parameters and project to 13B, experiments at truly large scale (100B+) remain future work due to resource constraints.
\end{enumerate}
