% DES-LOC: Desynced Low Communication Adaptive Optimizers for Foundation Models
% Reconstructed from 12473_DES_LOC_Desynced_Low_Com.pdf
% Claude-1 (M001-M025): Sections 1-7 main body LaTeX reconstruction
% Original: Published as a conference paper at ICLR 2026

\documentclass{article}
\usepackage{iclr2026_conference}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{multirow}

\title{DES-LOC: Desynced Low Communication Adaptive Optimizers for Foundation Models}

\author{
Alex Iacob$^{\dagger,1,2}$ \quad Lorenzo Sani$^{1,2}$ \quad Mher Safaryan$^{3}$ \quad Andrej Jovanovi\'{c}$^{*,1}$ \\
William F.\ Shen$^{1}$ \quad Xinchi Qiu$^{1}$ \quad Samuel Horv\'{a}th$^{*,5}$ \\
Preslav Aleksandrov$^{1,6}$ \quad Paris Giampouras$^{*,4}$ \quad Meghdad Kurmanji$^{*,1}$ \\
Nicholas D.\ Lane$^{1,2}$ \\
\\
$^{1}$ University of Cambridge \quad $^{2}$ Flower Labs \quad $^{3}$ ISTA \\
$^{4}$ University of Warwick \quad $^{5}$ MBZUAI \quad $^{6}$ IMEC \\
$^{\dagger}$\texttt{aai30@cam.ac.uk}; $^{*}$ Equal contributions
}

\begin{document}
\maketitle

\begin{abstract}
Scaling foundation model training with Distributed Data Parallel (DDP) methods
is bandwidth-limited. Existing infrequent communication methods like Local
SGD were designed to synchronize model parameters only and cannot be trivially
applied to adaptive optimizers due to additional optimizer states. Heuristic approaches that keep states local or reset them lack guarantees and can be unstable in
compute-efficient batch regimes; conversely, Local Adam synchronizes all states
uniformly and is provably convergent but triples communication costs. We propose
Desynced Low Communication Adaptive Optimizers (DES-LOC), a family of optimizers assigning independent synchronization periods to parameters and momenta,
enabling lower communication costs while preserving convergence. Our theoretical
analysis shows that while parameter synchronization dominates the asymptotic rate
in-expectation, high-probability convergence guarantees require at least infrequent
synchronization of the second momentum. Furthermore, we prove that more frequent momentum sync permits larger stable step sizes. Experiments on language
models of up to 1.7B show that DES-LOC can communicate $170\times$ less than DDP
and $2\times$ less than the previous state-of-the-art Local Adam, enabling $1.3$--$2.1\times$
wall-clock speedups over DDP for 1--13B models on 100Gb/s links. Furthermore,
unlike previous heuristic methods, DES-LOC is robust to worker failures offering
a scalable, efficient, and fault-tolerant solution for foundation model training.
\end{abstract}

%=============================================================================
\section{Introduction}
\label{sec:introduction}
%=============================================================================

Training foundation models requires distributing optimization across workers for improved memory
and compute. However, frequent gradient communication in standard Distributed Data Parallelism
(DDP)~\citep{li2020pytorch} increases networking costs and limits scalability. Early works like Local
SGD~\citep{stich2019local} and FedAvg~\citep{mcmahan2017communication} reduced this overhead by synchronizing
infrequently, averaging parameters only after $K \gg 1$ local steps, instead of gradients at every step.
However, modern foundation model training, e.g., Large Language Models~\citep{dubey2024llama}, uses
adaptive optimizers~\citep{kingma2015adam} which require additional momenta.

Some extensions of Local SGD to adaptive optimizers~\citep{sani2025photon,douillard2023diloco}
only average model parameters, which poses challenges. First, they lack convergence guarantees.
Second, keeping momenta local~\citep{douillard2023diloco} accumulates noisy small-batch gradients and
provides no means to initialize workers. This makes them unsuitable for failure-prone environments.
Third, re-initializing momenta~\citep{sani2024future,sani2025photon} destabilizes training.

Local Adam~\citep{cheng2025convergence} addresses these challenges, proving periodic synchronization
can converge faster than standard Adam with DDP, and remain robust to the addition of
new workers. However, it requires synchronizing momenta alongside model parameters, tripling
communication costs compared to Local SGD. Hence, we aim to answer the following question:

\begin{center}
\emph{Can independently syncing parameters and momenta improve communication
efficiency for adaptive optimizers while maintaining convergence and robustness?}
\end{center}

As a result of our inquiry, we propose a new optimizer family, Desynced Low Communication
Adaptive Optimizers (DES-LOC), which sets independent synchronization frequencies for parameters
and momenta. This reduces communication overhead by synchronizing momenta less frequently.

\paragraph{Contributions:}
\begin{enumerate}
    \item \textbf{Provable convergence.} We prove convergence (see Section~\ref{sec:convergence}) for DES-LOC under: non-convex
    objectives when using SGD with momentum (SGDM), and weakly convex objectives when using Adam. Our theory indicates a higher momentum sync frequency enables larger step sizes.
    Furthermore, high-probability bounds demand momenta be synced with finite period for $\beta_2 < 1.0$.

    \item \textbf{Communication reduction.} We empirically show that parameters require more frequent sync than
    momenta, and that less frequent momentum sync reduces communication costs ($2\times$ vs Local
    Adam, $170\times$ vs DDP), leading to $1.3$--$2.1\times$ reductions in training time over DDP on our hardware.

    \item \textbf{Scalability to large models.} We validate DES-LOC at billion-scale language model training,
    demonstrating competitive ICL performance against both Local Adam and DDP.

    \item \textbf{Hardware robustness.} Unlike previous heuristic methods, DES-LOC avoids persistent local states,
    enabling it to seamlessly integrate new workers to support environments prone to system failures.
\end{enumerate}

%=============================================================================
\section{Desynced Low Communication Adaptive Optimizers (DES-LOC)}
\label{sec:desloc}
%=============================================================================

We start by characterizing the relation between the rate of change of optimizer states and Local
Adam, and how these can be leveraged to lower the communication cost. Consider the Adam update:
\begin{equation}
u_t = \beta_1 u_{t-1} + (1 - \beta_1) g_t \quad \text{and} \quad v_t = \beta_2 v_{t-1} + (1 - \beta_2) g_t \odot g_t.
\end{equation}

For Local Adam, convergence is contingent on $\beta_2$ satisfying $1 - \beta_2 = \tilde{O}(K^{-3/2} R^{-1/2})$~\citep{cheng2025convergence} where $K$ is the number of local steps and $R$ the total communication rounds. Large
$K$ or $R$ implies $\beta_2 \to 1$, and conversely larger $\beta_2$ permits higher $K$ or $R$.

A useful summary measure is the number of steps until a state's weight decays to a fraction $\psi$,
$\tau_\psi(\beta) = \frac{\ln \psi}{\ln \beta}$. Following~\citet{pagliardini2025ademamix}, we use the half-life $\tau_{0.5}$ as our primary measure,
omitting $\beta$ when clear. For typical values of $\beta$, we have $\tau_{0.5}(0.95) \approx 13.5$~\citep{allal2025smollm2},
$\tau_{0.5}(0.999) \approx 692.8$~\citep{kingma2015adam}, and $\tau_{0.5}(0.9999) \approx 6931$~\citep{taniguchi2024adopt}.

Intuitively, larger half-lives imply synchronizing gradients over longer horizons as the optimizer
is less sensitive to new gradients; choosing $\beta = 0$ ignores all previous momenta, whereas $\beta \to 1$
progressively attenuates signal from the current gradient.

While the half-life captures the horizon for which an optimizer state remains relevant to model
updates, it provides no information on its absolute rate of change. With coordinate-wise clipping,
each gradient component satisfies $|(g_t)_i| \leq \rho$. Unrolling Adam's recursions over $K$ local steps gives
the follow relation: $u_{t+K} = \beta_1^K u_t + (1 - \beta_1) \sum_{k=0}^{K-1} \beta_1^k g_{t+K-1-k}$ and its second moment analogue.
Since $|g_{t,i}| \leq \rho$ and $|(g_t \odot g_t)_i| \leq \rho^2$, the maximal $\ell_\infty$ drift of each moment is (see Section~\ref{sec:drift_derivation}):
\begin{align}
\|u_{t+K} - u_t\|_\infty &\leq 2\rho(1 - \beta_1^K), \label{eq:drift_u} \\
\|v_{t+K} - v_t\|_\infty &\leq 2\rho^2(1 - \beta_2^K). \label{eq:drift_v}
\end{align}

From the above, large $\beta$ values and small clip bounds $\rho$, a common practice in foundation model
training~\citep{brown2020language,scao2022bloom}, limit the absolute changes in optimizer states. We can
construct similar reasoning for other optimizers~\citep{sutskever2013importance,taniguchi2024adopt}, and
norm-based clipping~\citep{pascanu2013difficulty,brown2020language}. From the above, the half-life of an
optimizer state should inform its synchronization frequency. For example, if $\tau_{0.5}(0.95) \approx 13.5$ and
$K = 256$, synchronization only affects few initial local steps. Over the course of the local training,
the impact of the synchronised optimizer state shall decay to $0$ given Equations~\ref{eq:drift_u} and~\ref{eq:drift_v}. Conversely,
if $K = 16$, synchronization approximately matches the half-life, strongly influencing local updates.

%-----------------------------------------------------------------------------
\subsection{DES-LOC Algorithm}
\label{sec:desloc_algorithm}
%-----------------------------------------------------------------------------

\begin{algorithm}[t]
\caption{DES-LOC}
\label{alg:desloc}
\begin{algorithmic}[1]
\REQUIRE Model tensors, update functions, hyper-parameters
\STATE $x_0 \in \mathbb{R}^d$, $\{s^j_{-1}\}_{j=1}^N \in (\mathbb{R}^d)^N$ \COMMENT{initial parameter vector, the initial $N$ optimizer states}
\STATE $\{\text{UPDATE}^j\}_{j=1}^N : (\mathbb{R}^d \times \mathbb{R}^d \to \mathbb{R}^d)^N$ \COMMENT{updates optimizer state $j$ from its previous state and the gradient}
\STATE $\text{OPT} : \mathbb{R}^d \times \mathbb{R}^d \times \mathbb{R}_+ \times (\mathbb{R}^d)^N \to \mathbb{R}^d$ \COMMENT{update params from all worker models}
\STATE $\text{SERVEROPT} : \mathbb{R}^d \to \mathbb{R}^d$ \COMMENT{update params using an abstract outer optimizer}
\STATE $\rho \in \mathbb{R}_+$, $\{\eta_t\}_{t=0}^{T-1} \in (\mathbb{R}_+)^{T-1}$ \COMMENT{clipping radius, learning-rate schedule}
\STATE $T, M \in \mathbb{N}_+$ \COMMENT{total optimization steps and number of workers}
\STATE $K_x \in \mathbb{N}_+$, $\{K_j\}_{j=1}^N \in (\mathbb{N}_+)^N$ \COMMENT{communication periods (steps)}
\ENSURE $x_T$, $\{s^j_{T-1}\}_{j=1}^N$
\STATE \textbf{for each worker} $m$: $x^m_0 \gets x_0$, $s^{j,m}_{-1} \gets s^j_{-1}$ \COMMENT{local init}
\FOR{$t = 0, \ldots, T-1$} \COMMENT{training loop}
    \FOR{all workers $m = 0, \ldots, M-1$ \textbf{in parallel}}
        \STATE $g^m_t \gets \nabla F(x^m_t; \xi^m_t)$ \COMMENT{stochastic grad}
        \STATE $\hat{g}^m_t \gets \text{clip}(g^m_t, \rho)$ \COMMENT{per-coordinate clipping}
        \FOR{$j = 1$ \TO $N$}
            \IF{$t \bmod K_j = 0$} \COMMENT{sync $s^j$}
                \STATE $s^{j,m}_t \gets \text{UPDATE}^j\left(\mathbb{E}_m[s^{j}_{t-1}], \hat{g}^m_t\right)$
            \ELSE
                \STATE $s^{j,m}_t \gets \text{UPDATE}^j\left(s^{j,m}_{t-1}, \hat{g}^m_t\right)$
            \ENDIF
        \ENDFOR
        \IF{$t \bmod K_x = 0$} \COMMENT{sync $x$}
            \STATE $x^m_{t+1} \gets \text{OPT}\left(\text{SERVEROPT}(\mathbb{E}_m[x^m_t]), \hat{g}^m_t, \eta_t, \{s^{j,m}_t\}_{j=1}^N\right)$
        \ELSE
            \STATE $x^m_{t+1} \gets \text{OPT}\left(x^m_t, \hat{g}^m_t, \eta_t, \{s^{j,m}_t\}_{j=1}^N\right)$
        \ENDIF
    \ENDFOR
\ENDFOR
\end{algorithmic}
\end{algorithm}

Motivated by the above insights, we formalize Desynced Low Communication Adaptive Optimizers
as a family of optimizers offering the same convergence and robustness as Local Adam but with
significantly lower communication costs. Our approach applies generically to adaptive optimizers
parameterized by $\text{OPT} : (\mathbb{R}^d, \mathbb{R}^d, \mathbb{R}_{>0}, \{\mathbb{R}^d\}^N) \to \mathbb{R}^d$, with $N$ optimizer states $\{s^j_{-1}\}_{j=1}^N \subset \mathbb{R}^d$,
each updated by $\text{UPDATE}^j : (\mathbb{R}^d, \mathbb{R}^d) \to \mathbb{R}^d$. Coordinate-wise clipping is defined as $[\text{clip}(X, \rho)]_i = \text{sgn}(X_i) \cdot \min\{|X_i|, \rho\}$. To ensure that our method is provably convergent, SERVEROPT is that of
FedAvg~\citep{mcmahan2017communication}. However, our algorithm directly extends to the larger FedOpt~\citep{reddi2021adaptive} framework, which we discuss in Section~\ref{sec:fedopt}.

We focus our analysis on SGDM and Adam. As shown in Algorithm~\ref{alg:desloc}, DES-LOC synchronizes
parameters $x \in \mathbb{R}^d$ and optimizer states $\{s^j\}_{j=1}^N$ at state-specific intervals $K_x, \{K_j\}_{j=1}^N \in \mathbb{N}_+$.
Setting $N = 2$, $s^1_t = u_t$, $s^2_t = v_t$, and using update rules $\text{UPDATE}^1$, $\text{UPDATE}^2$ based on the Adam
update rules above yields DES-LOC-Adam (see Algorithm~\ref{alg:desloc_adam}).

\paragraph{Toy Example} To highlight DES-LOC's practical benefit, Fig.~\ref{fig:toy} illustrates a scenario where DES-LOC
and Local Adam converge under noisy gradients, while prior heuristic methods~\citep{douillard2023diloco,sani2025photon,iacob2025dept,sani2024future} fail.

%=============================================================================
\section{Convergence Guarantees for DES-LOC}
\label{sec:convergence}
%=============================================================================

This section provides theoretical support for the proposed DES-LOC approach. We focus on a version
of the Adam optimizer that uses only a single momentum state. Extensions to the full Adam optimizer
with both momenta are available in Section~\ref{sec:adam_extension} with high-probability bounds shown in Section~\ref{sec:high_prob}.

Formally, we consider the following optimization problem:
\begin{equation}
\min_{x \in \mathbb{R}^d} f(x) := \frac{1}{M} \sum_{m=1}^{M} f_m(x), \quad \text{with } f_m(x) = \mathbb{E}_{\xi \sim \mathcal{D}_m}[F_m(x; \xi)].
\label{eq:objective}
\end{equation}

In this setup, all $M$ machines collaboratively minimize the objective in~\eqref{eq:objective}. Generally, we assume
each machine $m$ has access to only dataset $\mathcal{D}_m$, which can differ from device to device. This recovers
the homogeneous distribution case when all machines have the same dataset $\mathcal{D}_1 = \mathcal{D}_2 = \cdots = \mathcal{D}_M$
and minimize the same loss $f_1(x) = f_2(x) = \cdots = f_m(x) = f(x)$. We assume each machine $m$
computes mini-batch stochastic gradients corresponding to randomly selected samples $\xi \sim \mathcal{D}_m$ from
dataset $\mathcal{D}_m$. We further use the following technical assumptions on the problem structure.

\begin{assumption}[Lower bound and smoothness]
\label{asm:smooth}
The overall loss function $f: \mathbb{R}^d \to \mathbb{R}$ is lower bounded by some $f^* \in \mathbb{R}$ and all local loss functions $f_m$ are $L$-smooth:
\begin{equation}
\|\nabla f_m(x) - \nabla f_m(y)\| \leq L\|x - y\|, \quad \text{for any } x, y \in \mathbb{R}^d.
\end{equation}
\end{assumption}

\begin{assumption}[Unbiased noise with bounded stochastic variance]
\label{asm:noise}
The stochastic gradient $g^m$ of local loss function $f_m$ computed by machine $m$ is unbiased and the noise has bounded variance:
\begin{equation}
\mathbb{E}[g^m] = \nabla f_m(x), \quad \mathbb{E}[\|g^m_t - \nabla f_m(x)\|^2] \leq \sigma^2, \quad \text{for any } x \in \mathbb{R}^d.
\end{equation}
\end{assumption}

\begin{assumption}[Bounded heterogeneity]
\label{asm:heterogeneity}
For any $x \in \mathbb{R}^d$, the heterogeneity is bounded by
\begin{equation}
\frac{1}{M} \sum_{m=1}^{M} \|\nabla f_m(x)\|^2 \leq G^2 + B^2 \|\nabla f(x)\|^2.
\end{equation}
\end{assumption}

All three assumptions are standard and widely used in the convergence analysis of optimization
algorithms~\citep{yu2019linear,karimireddy2020scaffold,wang2021field,yuan2022revisiting}.
Note that the bounded heterogeneity condition recovers the homogeneous case when $G^2 = 0$ and
$B^2 = 1$. To facilitate the technical presentation of the analysis, we view model and optimizer
state synchronizations through assigning probabilities to each averaging event. Particularly, instead
of averaging model parameters every $K_x$ steps (i.e., $t \bmod K_x = 0$), we average with probability
$p_x = \frac{1}{K_x}$, which are statistically equivalent. In the following theorem, we provide convergence rate
of SGDM optimizer under such probabilistic and decoupled synchronization:

\begin{theorem}
\label{thm:sgdm}
Let Assumptions~\ref{asm:smooth},~\ref{asm:noise} and~\ref{asm:heterogeneity} hold. Then, choosing the step size $\eta = \min(\eta_0, \frac{1}{\sqrt{T}})$ with
\begin{equation}
\eta_0 \stackrel{\text{def}}{=} \frac{1}{4L} \min\left(1 - \beta, \frac{1}{\sqrt{\psi \max(1, B^2 - 1)}}\right), \quad \text{where } \psi \stackrel{\text{def}}{=} \frac{4(1-p_x)}{p_x^2} \cdot \frac{(1-\beta)(1-p_u)}{6(1-(1-p_u)\beta)},
\label{eq:step_size}
\end{equation}
the average iterates $\bar{x}_t = \mathbb{E}_m[x^m_t]$ of DES-LOC-SGDM converge with the following rate:
\begin{equation}
\frac{1}{T} \sum_{t=0}^{T-1} \mathbb{E}\|\nabla f(\bar{x}_t)\|^2 \leq \frac{4}{\sqrt{T}} \left(f(x_0) - f^* + \frac{L\sigma^2}{2M}\right) + O\left(\frac{1+\psi}{T}\right).
\label{eq:rate}
\end{equation}
\end{theorem}

We now discuss the convergence result and its implications. The obtained rate~\eqref{eq:rate} is asymptotically
optimal for this setup~\citep{arjevani2023lower}. Notably, the leading term $O(\frac{1}{\sqrt{T}})$ is unaffected by the
number of local steps. Interestingly, probabilities $p_x$, $p_u$, and the momentum parameter $\beta$ appear in
the higher-order term $O(\frac{1}{T})$, and thus have a limited impact on asymptotic convergence speed.

Regarding state synchronization, it is evident from~\eqref{eq:step_size} that model synchronization has a greater
impact on convergence due to the dependence $\psi = O(\frac{1}{p_x^2})$. With vanishing $p_x$, the $\psi$ term becomes
unbounded and breaks the rate. For optimizer states, it seems that momentum averaging can be
turned off ($p_u = 0$) without affecting the asymptotic behavior of the rate. Setting $p_x = 1$ and $p_u = 0$
recovers standard mini-batch SGDM~\citep{liu2020improved}. However, the $\psi$ term also appears in the step-size restriction~\eqref{eq:step_size}. As $p_u \to 0$, $\frac{(1-p_u)}{1-(1-p_u)\beta} \to \frac{1}{1-\beta}$. This imposes the most severe restriction on the
learning rate $\eta_0$ since $\eta_0 \propto \frac{1}{\sqrt{\psi}}$, as $\psi$ is maximized. This theory shows that increasing the frequency
$p_u$ of momentum averaging---while not changing the asymptotic rate---allows for a larger step size,
potentially leading to faster convergence in practice. This theory justifies that momentum states
can be synchronized less frequently than parameters and that more averaging improves convergence
by supporting larger step sizes. Furthermore, our high probability analysis of DES-LOC-Adam in
Section~\ref{sec:high_prob} shows that the sync frequency of momenta must be finite for $\beta_2 < 1.0$.

%=============================================================================
\section{Experimental Design}
\label{sec:experimental_design}
%=============================================================================

Our experimental setup addresses the following research questions:
\begin{description}
\item[RQ1] Do theoretical rates of change predict the empirical evolution of optimizer states?
\item[RQ2] How does the synchronization frequency of a model/optimizer state impact performance?
\item[RQ3] To what extent can DES-LOC cut communication w.r.t.\ Local Adam in practical scenarios?
\item[RQ4] How does DES-LOC scale with increasing model size and longer training horizons?
\item[RQ5] How does DES-LOC perform when using a Nesterov outer optimizer?
\item[RQ6] Is DES-LOC compatible with non-Adam inner optimizers such as Muon?
\end{description}

%-----------------------------------------------------------------------------
\subsection{Experimental Setup}
\label{sec:experimental_setup}
%-----------------------------------------------------------------------------

\paragraph{Models and data.} We train a 135M-parameter GPT-style model (arch.\ in Table~\ref{tab:arch}) with sequence
length 2048. We distinguish worker batch size $B_w$ from global $B = \sum_{w=0}^{M-1} B_w$~\citep{sani2025photon}. A
2M token global batch is split across $M = 4$ workers sampling IID from SmolLM2~\citep{allal2025smollm2}:
70\% Fineweb-Edu~\citep{penedo2024fineweb}, 10\% Cosmopedia~\citep{benallal2024cosmopedia}, 10\%
Python-Edu, 5\% FineMath 4+, and 5\% Infi-WebMath 4+. The 135M model trains for
6.4B tokens ($2.4\times$ compute-optimal~\citep{hoffmann2022training}). For RQ4, we scale to a 1.7B model for
40B tokens ($2\times$ compute-optimal)~\citep{sardana2024beyond}. We show Non-IID data results in Fig.~\ref{fig:noniid}.

\paragraph{Optimizers.} We use Adam~\citep{kingma2015adam} (results in Section~\ref{sec:complementary}) and its variant
ADOPT~\citep{taniguchi2024adopt}, which modifies the update to guarantee convergence for any $\beta_2$.
For the 135M model, we grid-search $(\beta_1, \beta_2, \eta)$ under DDP; the 1.7B model uses hyperparameters
from~\citet{allal2025smollm2,taniguchi2024adopt}. Learning rates use the WSD schedule~\citep{hagele2024scaling,allal2025smollm2}. We favor ADOPT ($\beta_2 = 0.9999$) in high-$\beta$ regimes where Adam is unstable.
We also ablate the outer optimizer, comparing FedAvg with a Nesterov optimizer~\citep{reddi2021adaptive,douillard2023diloco,charles2025communication} on a 700M model trained on 40B tokens.

\paragraph{Baselines.} We compare DES-LOC with: (i) synchronous DDP; (ii) Local Adam/ADOPT; (iii)
FAVG+OPT (persistent states~\citep{sani2025photon,douillard2023diloco}); and (iv) FAVG$-$OPT
(reset states~\citep{sani2024future,iacob2025dept}). Persistent-state FedAvg is DES-LOC with
$K_u, K_v = \infty$, an upper bound on comms efficiency. DDP is an upper bound on ML performance.

\paragraph{Metrics.} We evaluate models by (i) perplexity and (ii) per-worker asymptotic communication cost
assuming a bandwidth-optimal Ring-AllReduce~\citep{sergeev2018horovod} algorithm scaling
linearly with model size. For the 1.7B model, we report standard in-context-learning (ICL) benchmarks~\citep{brown2020language}. We use a zero-shot setting for ICL tasks unless stated otherwise following~\citet{allal2025smollm2} and report the best performing communication-efficient method in blue with the
best-performing overall in bold.

%=============================================================================
\section{Evaluation}
\label{sec:evaluation}
%=============================================================================

Our results show optimizer states change at different rates (Section~\ref{sec:rq1}), forming a clear synchronization hierarchy (Section~\ref{sec:rq2}). DES-LOC reduces communication $2\times$ vs.\ Local Adam (Section~\ref{sec:rq3})
while converging robustly with adding workers and scaling effectively to large models (Section~\ref{sec:rq4}).

%-----------------------------------------------------------------------------
\subsection{Higher $\beta$ Optimizer States Have Slower Empirical Rates of Change (RQ1)}
\label{sec:rq1}
%-----------------------------------------------------------------------------

Figure~\ref{fig:rates_of_change} shows that relative rates of change for the two momenta in Local ADOPT/Adam scale
with their decay rates under gradient clipping ($\rho = 1$). Supported by our theoretical discussions on
momenta half-lives (Section~\ref{sec:desloc}), the second momentum evolves substantially slower than the first at
high-$\beta_2$. For Local Adam, the second momentum remains slower even when $\beta_2 \approx \beta_1$, potentially
because gradient variance~\citep{kingma2015adam} evolves slower than the first momentum.

\paragraph{Takeaway:} As discussed in Sections~\ref{sec:desloc} and~\ref{sec:convergence}, when $\beta_1 \ll \beta_2$, the second momentum evolves slower
than the first, proportional to half-life ratio of the two $\frac{\tau_{0.5}(\beta_2)}{\tau_{0.5}(\beta_1)} = \frac{\ln(\beta_1)}{\ln(\beta_2)}$.

%-----------------------------------------------------------------------------
\subsection{Parameters Require Frequent Sync, Momenta Sync Proportional to $\beta$ (RQ2)}
\label{sec:rq2}
%-----------------------------------------------------------------------------

Figure~\ref{fig:sync_freq} evaluates the effect of independently varying synchronization periods ($K_x, K_u, K_v$) for
parameters and optimizer states. We consider two baseline periods ($K_b = 16, 256$), chosen based
on the fastest state's half-life ($\tau_{0.5}(0.95) \approx 13.5$). Frequent parameter synchronization ($K_x$) is
crucial for performance, while synchronizing momenta ($K_u, K_v$) significantly impacts training only
if their half-lives align with the base frequency $K_b$. Otherwise, synchronization frequency primarily
influences communication costs rather than model quality.

\paragraph{Takeaway:} Parameter synchronization frequency ($K_x$) strongly impacts performance, motivated
by the leading term in theoretical bounds (Section~\ref{sec:convergence}). Momentum synchronization periods matter
empirically only when chosen near their half-lives, consistent with Sections~\ref{sec:desloc} and~\ref{sec:convergence}.

%-----------------------------------------------------------------------------
\subsection{DES-LOC Brings $2\times$ Communication Reductions over Local Adam (RQ3)}
\label{sec:rq3}
%-----------------------------------------------------------------------------

As shown in Figure~\ref{fig:comms}, DES-LOC halves communication versus Local Adam~\citep{cheng2025convergence}
with matching perplexity by syncing momenta less frequently ($K_u = 3K_x$, $K_v = 6K_x$),
exploiting the second-momentum's lower sensitivity to sync frequency (Fig.~\ref{fig:sync_freq}). This yields a $1.37\times$
speedup over DDP and $1.1\times$ over Local Adam at $K_x = 16$ on 4 H100s. At $K_x = 256$, the speedup
over DDP increases to $1.47\times$ and both DES-LOC and Local Adam saturate throughput.

As shown in Figure~\ref{fig:comms}, DES-LOC effectively initializes new workers and outperforms the straightforward approach of ad-hoc averaging from checkpoints---whose insufficiency is detailed in Section~\ref{sec:checkpoint}.

\paragraph{Takeaway:} DES-LOC achieves a $2\times$ communication reduction over Local Adam by leveraging two
insights: optimizer-state sync matters less than parameter sync, and slower-changing states (high $\beta_2$)
can sync less often. By eventually syncing all optimizer states, DES-LOC matches the robustness of
Local Adam with $K = \max(K_x, K_u, K_v)$ when adding new workers/responding to system failures.

%-----------------------------------------------------------------------------
\subsection{DES-LOC is Suitable for Large-Scale Training (RQ4)}
\label{sec:rq4}
%-----------------------------------------------------------------------------

Figure~\ref{fig:billion} shows that DES-LOC reliably scales to 1B models and very infrequent communication ($K_x = 256$). Evaluating the billion-scale models on the ICL tasks (Table~\ref{tab:icl}), DES-LOC is
competitive with all baselines while reducing communication versus Local Adam and DDP. The
heuristic baseline~\citep{sani2025photon} suffers training instabilities (Fig.~\ref{fig:billion}b) potentially impacting
downstream performance (Table~\ref{tab:icl}) and underscoring the advantage of DES-LOC's training stability.

Our method's reduced communication costs result in a $\approx 2.2\times$ training speedup over DDP (Figure~\ref{fig:billion}).

\begin{table}[t]
\centering
\caption{Our billion-scale model trained with DES-LOC matches or surpasses the ICL performance of models trained with Local Adam and FAVG+OPT, approaching DDP performance.}
\label{tab:icl}
\begin{tabular}{lcccc|c}
\toprule
& Arc Challenge & Arc Easy & PIQA & HellaSwag & Avg \\
\midrule
DES-LOC      & 31.8 & 59.0 & 70.7 & 44.9 & 51.6 \\
Local Adam   & 31.9 & 59.0 & 70.6 & 45.8 & 51.8 \\
FAVG+OPT     & 30.1 & 58.0 & 70.0 & 44.8 & 50.7 \\
\midrule
DDP          & 33.8 & 62.5 & 71.1 & 47.8 & 53.8 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{Wall-clock time (days) for 1B--13B models to reach $2\times$ compute-optimal tokens.}
\label{tab:wallclock}
\begin{tabular}{l|cc|cc|cc}
\toprule
& \multicolumn{2}{c|}{1B Model} & \multicolumn{2}{c|}{7B Model} & \multicolumn{2}{c}{13B Model} \\
$K_x$ & 16 & 256 & 16 & 256 & 16 & 256 \\
\midrule
DDP          & $1.41$ & $1.41$ & $38.74$ & $38.74$ & $175.50$ & $175.50$ \\
\midrule
FAVG+OPT     & $0.80$ & $0.63$ & $28.52$ & $24.01$ & $100.21$ & $82.46$ \\
Local Adam   & $0.96$ & $0.64$ & $31.46$ & $24.18$ & $116.10$ & $83.34$ \\
DES-LOC      & $0.81$ & $0.63$ & $28.80$ & $24.06$ & $102.03$ & $82.68$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Takeaway:} DES-LOC enables efficient training of large-scale foundation models, especially at long
training horizons, with downstream ICL performance competitive with DDP. We recommend setting
$K_x$ for sufficient throughput based on bandwidth, then setting $K_u, K_v$ as constant multiples (e.g.,
$3\times$, $6\times$) or based on the half-life of their $\beta$ (see Section~\ref{sec:choosing_freq}).

%-----------------------------------------------------------------------------
\subsection{Nesterov as the Outer Optimizer (RQ5)}
\label{sec:rq5}
%-----------------------------------------------------------------------------

We ablate the outer optimizer for DES-LOC on a 700M parameter model, comparing averaging to a
Nesterov optimizer with momentum of 0.9, outer learning rate of 1.0 tuned following~\citet{charles2025communication}. The experiment ran on 4 H100s and used a medium-synchronization regime ($K_x = 32$, $K_u =
3K_x$, $K_v = 6K_x$) where models are initialised from 2048-step DDP checkpoints, following~\citet{charles2025communication}.

\paragraph{Takeaway:} Frequent synchronization ($K_x = 32$) allows DES-LOC to reach within 1\% of the perplexity of DDP. Furthermore, using a Nesterov outer optimizer improves performance over averaging,
while preserving practical benefits like effective worker initialization and reduced optimization noise.

%-----------------------------------------------------------------------------
\subsection{Muon as the Inner Optimizer (RQ6)}
\label{sec:rq6}
%-----------------------------------------------------------------------------

To assess the versatility of DES-LOC beyond Adam-family optimizers, we integrate it with Muon~\citep{jordan2024muon}, a single-momentum optimizer that uses Newton-Schulz iterations for orthogonalization.
Since Muon preconditions the momentum term directly rather than tracking second-moment
estimates, the relevant synchronization periods reduce to parameters ($K_x$) and momentum ($K_u$). We
apply DES-LOC by synchronizing the Muon momentum buffer less frequently than the parameters,
setting $K_u = 3K_x$ with $K_x = 32$.

\paragraph{Takeaway:} DES-LOC is compatible with single-momentum optimizers such as Muon that rely
on Newton-Schulz preconditioning. By reducing the synchronization frequency of the momentum
buffer relative to parameters, DES-LOC maintains solution quality while significantly lowering
communication volume, demonstrating that our approach is inner-optimizer-agnostic.

%=============================================================================
\section{Related Work}
\label{sec:related}
%=============================================================================

In synchronous data-parallel training, workers exchange full gradients or parameters every iteration,
incurring communication costs linear in model size using Ring-AllReduce~\citep{sergeev2018horovod}.
When hardware is weakly connected or widely distributed, communication significantly
slows wall-clock training time~\citep{sani2025photon} as workers need to wait for synchronization to
finish. Federated Averaging (FedAvg)~\citep{mcmahan2017communication} and Local SGD~\citep{stich2019local}
reduce communication by performing $K$ local optimization steps before averaging parameters,
decreasing communication rounds by a factor of $K$.

Adam~\citep{kingma2015adam} is popular for pre-training as it scales to larger batches than SGD~\citep{kunstner2023noise,dubey2024llama}. It uses moving averages of gradients and their squares, however,
its convergence is not guaranteed as it requires $\beta_1 < \sqrt{\beta_2} < 1$, with large, problem-specific $\beta_2$~\citep{reddi2018convergence,zhang2022adam}. Local Adam~\citep{cheng2025convergence} reduces communication with local steps but requires syncing optimizer states, which triples the
communication cost relative to Local SGD/DDP, as sync costs scale with the number of states.

%=============================================================================
\section{Conclusion}
\label{sec:conclusion}
%=============================================================================

DES-LOC reconciles communication efficiency with rigorous convergence guarantees in distributed
adaptive optimization. By extending theory to the independent synchronization of Adam and SGDM
optimizer states, we empirically demonstrate convergence alongside $170\times$ and $2\times$ communication
reductions over DDP and prior state-of-the-art methods at billion-scale LLM training, even in environments prone to system failures. Our findings yield clear guidelines: (i) frequently synchronize
parameters, and (ii) synchronize optimizer states less often, proportional to their half-lives. These
insights open avenues for future research, including layer-wise synchronization, adaptive frequencies,
compressed updates, as well as emerging applications, such as worldwide cross-data center training
and collaborative training. As training workloads scale, we envision DES-LOC becoming the standard
for efficient, resilient foundation-model training in data centers and distributed environments.

\section*{Acknowledgments}
All costs for the computational resources used for this work were funded by Flower Labs, and the
research conducted by a team of researchers from Flower Labs, The University of Cambridge, The
Institute of Science and Technology of Austria, The University of Warwick, and Mohamed bin
Zayed University of Artificial Intelligence.

\bibliography{references}
\bibliographystyle{iclr2026_conference}

\end{document}
